---
title: "Introduction to blocking techniques and traditional record linkage"
author: "Rebecca C. Steorts"
date: |
  | July 10, 2019
  |
  |
output:
  beamer_presentation:
    keep_tex: no
    template: beamer.tex
  ioslides_presentation: default
  slidy_presentation: default
institute: |
  | Department of Statistical Science, affiliated faculty in Computer Science, 
  | Biostatistics and Bioinformatics, the information initiative at Duke (iiD) and
  | the Social Science Research Institute (SSRI) 
  | Duke University and U.S. Census Bureau
  | beka@stat.duke.edu
  | Population Dynamics and Health Program Workshop, University of Michigan
natbib: yes
fig_caption: yes
shortinstitute: bb222@stat.duke.edu
classoption: compress
---

# Blocking: Motivation

- Naively matching two files or finding duplicates within a file requires comparing all pairs of records.

\vspace{3mm}

- Infeasible for large files even when the comparisons are computationally inexpensive.

\vspace{3mm}

- The number of record pairs grows quadratically with the size of
the dataset
    * Two files with 5,000 records $\rightarrow$ 25,000,000 comparisons!

# What is blocking?

Technique to reduce the comparison space: 

- Filter out dissimilar record pairs that are extremely unlikely to be matches.
    * Perform record linkage only within blocks 


\vspace{3mm}

- Traditional blocking : compare record pairs that match on one or more keys.
    * Creates a partition of the data

\vspace{3mm}

- Record pairs that do not meet the blocking criteria are automatically classified as non-matches.


# Example: Traditional blocking

All-to-all record comparisons (left) versus partitioning records into
blocks by lastname initial and comparing records only within each partition (right).

![alt text](figures/noblocking_plot.pdf){ width=50% }
![alt text](figures/blocking_plot.pdf){ width=50% }


# Example: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(RecordLinkage)
data(RLdata500)
head(RLdata500)
```

# Continuation: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}

# Record pairs for comparison
choose(500,2)

# Blocking by last name initial  
last_init <- substr(RLdata500[,"lname_c1"], 1, 1)
head(last_init)
# Number of blocks
length(unique(last_init))

```
# Continuation: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}

# Number of records per block
tbl <- table(last_init)
head(tbl)

# Block sizes can vary a lot
summary(as.numeric(tbl))

```

# Continuation: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}

# Number of records pairs per block
sapply(tbl, choose, k=2)

# Reduction on comparison space
sum(sapply(tbl, choose, k=2))

```

# Blocking caveats

- Fields can be unreliable for many applications and blocking may miss large proportions of matches i.e. increased false negatives rates.

\vspace{3mm}

- The frequency distribution of the values in the fields used as blocking keys will affect the size of the blocks.

\vspace{3mm}

- Trade-off between block sizes: true matches being missed vs computational efficiency.

# How to choose the blocking key or keys

- Fields containing the fewest errors or missing values should be chosen as blocking variables e.g. clinical diagnosis in EHR.

\vspace{3mm}

- Understand the kind of errors that are unlikely for a certain field or a combination of them.

\vspace{3mm}

- More complex blocking schemes can be constructed using conjunctions.
    * Retain only pairs which agree on last name initial and zip code.

# Example: Voter Survey data

The Views of the Electorate Research (VOTER) Survey was conducted by the survey firm YouGov. 

\vspace{2mm}

- 8,000 adults (age 18+) with internet access took the survey on-line between November 29 and December 29, 2016.

\vspace{2mm}

- These respondents were originally interviewed by YouGov in 2011-2012.

\vspace{2mm}

- Barack Obama (Democrat) won in 2012 and Donald Trump (Republican) won in 2016.

\textcolor{blue}{https://www.voterstudygroup.org}




# Continuation: Voter Survey data

- Demographic variables
    * Year of birth (age) 
    * Gender
    * Race
    * State
    * Education level
    * Family income

\vspace{3mm}

- Party affiliation: democrat, republican, independent, other 

\vspace{3mm}

Which fields are reliable for blocking in this example?

# Continuation: Is race reliable?

\definecolor{LightCyan}{rgb}{0.88,1,1}
\definecolor{Gray}{gray}{0.9}
\begin{center}
\begin{tabular}{l|cc}

           &     2012 &  2016 \\ 
\hline \hline          
White      &     6244 &  6198 \\ 
Black      &       654 &  645 \\ 
Hispanic   &      400 &  397  \\ 
Mixed      &       160 &  186  \\ 
Other      &      137 &  167  \\ 
Asian      &       117 &  118  \\ 
Native American &   60 &   59 \\ 
Middle Eastern  &  10 &   12 \\ \hline \hline
\end{tabular}

\begin{tabular}{r|cccc}
      & White & Black & Mixed & Other \\ 
\hline \hline      
White & 6073  &   5   & {\cellcolor{LightCyan}46}    & 74 \\ 
Black &    4  & 627   & 10    & 10 \\ 
Mixed &  {\cellcolor{LightCyan}31}  &   6   & 100   &  8 \\ 
Other &   50  &   4   & 14    & 62 \\ \hline \hline
\end{tabular}
\end{center}

# Continuation: Is party affiliation reliable?

\begin{center}
\begin{tabular}{l|ccccc}
           & Democrat   & Indepen.  & Republican & Not sure & Other\\
\hline \hline            
  Democrat    &    2424   &       192   &      90   &    25  &   23 \\
  Indepen. &     263   &     1929    &    221    &   16  &  57 \\
  Republican  &      39   &      215    &   1881    &   11  &  60 \\
  Not sure    &      48   &       48    &     54    &   41  &   5 \\
  Other       &      17   &       46    &     34    &    2  &  41 \\
  \hline \hline
\end{tabular}
\end{center}


#  Record Linkage: Approximate matching

- Performing exact matching is very rare in practice since data is
rarely noise/error free.

\vspace{1mm}

- Matching that allows fields to only be similar rather than exact duplicates.

\vspace{1mm}

- Most large-scale applications employ some level of approximate match between fields.

\vspace{1mm}

- Techniques for direct matching include edit distance and hashing algorithms for string data.

# Example: Iterative deterministic linkage

\textcolor{blue}{Step 1:} two records must match on SSN and one of the following:

\vspace{1mm}

- First and last name.
- Last name, month of birth, and sex.
- First name, month of birth, and sex.

\vspace{1mm}

\textcolor{blue}{Step 2:} If SSN is missing or does not match, two records must match on last name, first name, month of birth, sex, and one of the following:

\vspace{1mm}

- Seven to eight digits of the SSN.
- Two or more of the following: year of birth, day of birth, middle initial, or date of death.

<!-- it would end up being an approximate matching that is a result of a two step deterministic matching -->

# Approximate String Matching: Similarity metrics 

Best for short strings such as person names:

- \textcolor{blue}{Levenshtein (edit) (1966)}: minimum number of substitutions required to transform one string into another e.g. A\textcolor{red}{d}a\textcolor{blue}{m} vs A\textcolor{blue}{l}a\textcolor{blue}{n} has a distance $L=2$, normalized as $1-\frac{L}{maxLength} = 0.5$ for similarity.

\vspace{1mm}

- \textcolor{blue}{Jaro-Winkler (1990):} The Jaro distance (1989) considers common characters and character transpositions. The JW similarity measure is:

\begin{center}
$ JW(A,B)= J(A,B) + \textcolor{red}{0.1p}(1-J(A,B)) $
\end{center}

where $p$ is the \# of the first four characters that agree exactly e.g. Adam vs Alan: p=1,  J= 0.67 and JW=0.7.

<!-- k-grams or k-shingles are better for longer or multi-word strings -->

# Example: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(RecordLinkage)
data(RLdata500)
```
```{r, eval=TRUE, echo=F, message=FALSE, warning=FALSE}
library(RecordLinkage)
data(RLdata500)
# Code to extract subset of duplicate records
RLdata500c <- RLdata500[,-c(2,4)]
cln <- table(identity.RLdata500)
iddup <- which(cln>1)-1
dup <- which(identity.RLdata500%in%iddup)
sub_dup <- RLdata500c[dup,]
oid <- order(identity.RLdata500[dup])
dup_set <- sub_dup[oid,]
tail(dup_set)
```

# Example: RLdata500

```{r, eval=TRUE, message=FALSE, warning=FALSE}
# Levenshtein similarity 
levenshteinSim("SCHUTE", "SCHULTE")
levenshteinSim("CHRISTA", "CHRISTAH")
# Jaro-Winkler similarity
jarowinkler(c("SCHUTE","CHRISTA"),
            c("SCHULTE","CHRISTAH"))
```

# Soundex algorithm


- Generates a code that represents the phonetic pronunciation of a word, helps identifying spelling variations of names. 

\vspace{2mm}

- The Soundex code for a name consists of a letter followed by three numerical digits: 
    * the letter is the first letter of the name,
    * the digits encode the remaining consonants.

\vspace{2mm}

- Consonants at a similar place of articulation share the same digit 
    * e.g. the labial consonants B, F, P and V are each encoded as the number 1.

# Example: Soundex algorithm
```{r, eval=TRUE, echo=F, message=FALSE, warning=FALSE}
tail(dup_set)
```
```{r, eval=TRUE, message=FALSE, warning=FALSE}
tail(soundex(dup_set$fname_c1))
tail(soundex(dup_set$lname_c1))
```

# Example: Soundex algorithm

```{r, eval=TRUE, echo=F, message=FALSE, warning=FALSE}
head(dup_set)
```
```{r, eval=TRUE, message=FALSE, warning=FALSE}
head(soundex(dup_set$lname_c1))
```

# Blocking by disjunctions 

- Produces overlapping blocks of the data.
    * \textcolor{blue}{Disjunction}: records match on field A or field B

\vspace{3mm}

- Using multiple keys to consider typographical
or measurement errors that would exclude true matches.
    * Blocking by last name initial or zip code
    
\begin{center}
\begin{tabular}{ l l l l }
 {\color{blue}\emph{1.}} & Mary Clain & 123 Oak St & 90210 \\
 {\color{blue}\emph{2.}} & Mary Klein & 123 Oak Street & 90210 \\
 {\color{blue}\emph{3.}} & Mary Klain & 123 Oak St & 50210 \\
\end{tabular}
\end{center}

- Reduction in false negative rates.

# Example: Blocking by disjunctions

```{r, eval=TRUE, message=FALSE, warning=FALSE}
# Two records must agree in either first name initial 
# or bith year to be compared.
# Only 2709 pairs instead of 124750!

rpairs <- compare.dedup(RLdata500c, 
blockfld = list(1, 3), #list with blocking fields
identity = identity.RLdata500)

tail(rpairs$pairs)
```

# Example: String comparison and blocking

```{r, eval=TRUE, message=FALSE, warning=FALSE}
rpairsfuzzy <- compare.dedup(RLdata500c, 
               phonetic = FALSE, blockfld = 3, 
               strcmp = TRUE, strcmpfun = jarowinkler)

tail(rpairsfuzzy$pairs)
```

# The Fellegi-Sunter approach (1969)

- Represent every pair of records using vector of features that describe similarity between individual record fields. 
    * Use string metrics for names and strings of numbers (Levenshtein or Jaro-Winkler).

\vspace{2mm}

- Place feature vectors for record pairs into three classes: matches $(M)$, nonmatches $(U)$, and possible matches.

\vspace{2mm}

- Let $P(\gamma|M)$ and $P(\gamma|U)$ be probabilities of observing a feature vector $\gamma$ for a matched and nonmatched pair, respectively.

# The Fellegi-Sunter approach (1969)

- Perform record-pair classification by calculating the ratio $w=(P(\gamma|M)/P(\gamma|U))$ for each candidate record pair.

\vspace{2mm}

- Establish two thresholds based on desired error levels to optimally separate the weight values for matches, possible matches, and nonmatches.

\vspace{2mm}

-  \textcolor{blue}{Note}: the quality of classification of the Fellegi-Sunter method relies strongly on reasonable estimations of $M$ and $U$ probabilities.

# Example: Blocking and Fellegi-Sunter

```{r, eval=TRUE, message=FALSE, warning=FALSE}

# tail(rpairs$pairs)
# Using comparison data blocking by first name initial
# and birth year
rpairs1 <- epiWeights(rpairs)

# Weights to compute thresholds for classification
rpairs1$Wdata[1:5]
```
# Example: Fellegi-Sunter

```{r, eval=FALSE, message=FALSE, warning=FALSE}
summary(rpairs1)
```

Weight distribution:

\begin{center}
\begin{tabular}{ccccc}
(0.35,0.4] & (0.4,0.45] & (0.45,0.5] & (0.55,0.6] & (0.6,0.65]\\      
      2    &     10     &       30   &      50    &      8 \\
(0.65,0.7] & (0.7,0.75] & (0.75,0.8] & (0.8,0.85] & (0.85,0.9] \\
     0     &     0     &        35  &        8   &       3 \\
\end{tabular}
\end{center}

# Example: Fellegi-Sunter

```{r, eval=FALSE, message=FALSE, warning=FALSE}
result <- epiClassify(rpairs1, 0.7)
summary(result)

alpha error: 0.080000 # False negative rate
beta error: 0.000000  # False positive rate
accuracy: 0.998523

Classification table:

           classification
true status    N    P    L
      FALSE 2659    0    0
      TRUE     4    0   46
```

# Summary

- \textcolor{blue}{Blocking}: reduce comparison space by choosing relatively noise free fields to match records 
    * Use conjunctions (and) create a partition of the data or disjunctions (or) to create overlapping blocks.
    
\vspace{2mm}

- \textcolor{blue}{Strings}: choose a string similarity metric to compare
record pairs within blocks.
    * Levenshtein or Jaro-Winkler.

\vspace{2mm}

- \textcolor{blue}{Record linkage}: use Fellegi-Sunter method to classify records as matches, posibble matches or nonmathches.


# References

\small

- Sariyar M. and Borg A. (2010), The RecordLinkage Package: Detecting
Errors in Data, The R Journal Vol. 2/2. \textcolor{blue}{Check big data functions of the package (>=1'000,000 record pairs).}

- Steorts et al. (2014) A Comparison of Blocking Methods for Record Linkage. In: Domingo-Ferrer J. (eds) Privacy in Statistical Databases. PSD 2014. Lecture Notes in Computer Science, vol 8744. Springer, Cham.

- Fellegi I.P., Sunter A.B. (1969), A Theory for Record Linkage, Journal of the American Statistical Association 64(328), pp. 1183–1210.

- Dusetzina et al. Linking Data for Health Services Research: A Framework and Instructional Guide. Rockville (MD): Agency for Healthcare Research and Quality (US) (2014). An Overview of Record Linkage Methods. https://www.ncbi.nlm.nih.gov/books/NBK253312/

- Entity Resolution and Information Quality, John R. Talburt, Elsevier (2011)

<!--
# Cluster-based Blocking

- Records agree on the blocking variables but are still very different 
    * e.g. two different people with the same lastname initial and gender

\vspace{2mm}

- The records in a cluster should be similar and therefore good candidate pairs for linkage.

\vspace{2mm}

- Methods to find clusters based on strings and cheap distance measures  
    * Threshold Nearest Neighbor 
    * K-Nearest Neighbor 
    * Canopies (fuzzy blocking)

# TNN and KNN

- Start with a single record as the base of the first cluster.

\vspace{2mm}

- Recursively add the nearest neighbors of records to the cluster until the distance to the nearest neighbor exceeds some threshold $t$.

\vspace{2mm}

- Choose one of the remaining records to start the next cluster.

\vspace{2mm}

- For KNN, ensure that each cluster has at least k records.

# Canopies

- Canopies is not strictly a blocking method $\rightarrow$ fuzzy blocking

\vspace{1mm}

- Records can be assigned to multiple blocks $\rightarrow$ overlapping clusters not a partition of the data.
    * Randomly pick a record to be the base of the first canopy
    * Records within distance $t_1$ are grouped into that canopy 
    * Remove records within distance $t_2 \leq t_1$ of the base
    * Pick another new record to start a second canopy


# Drawbacks

- Rough distance measures for complicated high-dimensional records are non-trivial.

\vspace{2mm}

- Requires actually performing all or nearly all comparisons to compute  
pairwise distances.

\vspace{2mm}

- Note that performing traditional blocking as a first step can reduce the space considerably.

[[I think I'd like to talk about hierarchical clustering and kmeans and show an illustration with the package, don't know if instead of TNN and KNN or added]]
-->