---
title: "Introduction to Bayesian record linkage "
author: Brenda Betancourt and Andee Kaplan
institute: |
    | Duke University
    | Department of Statistical Science
shortinstitute: Duke University
date: |
  | February 8, 2018
  |
  | Slides available at <http://bit.ly/cimat-bayes>
  |
output: 
  beamer_presentation:
    keep_tex: false
    template: beamer.tex
fig_caption: true
classoption: compress
natbib: true
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
theme_set(theme_bw(base_family = "serif", base_size = 30))
```

# What is "Bayesian"?

1. Setting up a *full probability model* -- a joint probability distribution for all observable and unobservable quantities
    \begin{align*}
    p(\boldsymbol x | \boldsymbol \theta) &- \text{ likelihood} \\
    p(\boldsymbol \theta) &- \text{ prior}
    \end{align*}
2. Conditioning on observed data -- calculating and interpreting the appropriate *posterior distribution*
    $$
    p(\boldsymbol \theta | \boldsymbol x) = \frac{p(\boldsymbol x, \boldsymbol \theta)}{p(\boldsymbol x)} = \frac{p(\boldsymbol x|\boldsymbol \theta)p(\boldsymbol \theta)}{p(\boldsymbol x)} \propto p(\boldsymbol x|\boldsymbol \theta)p(\boldsymbol \theta)
    $$

# Why Bayes?

A Bayesian framework is suitable to solve the following
problems:

\vspace{1mm}

- Exact computation of the probability that each pair is a match,
conditional on the observed data. 
    * Results conditioning on observed events are more directly interpretable than those obtained by conditioning on unobservable hypotheses.

\vspace{2mm}

- Propagating linkage error as an added component of uncertainty in the estimation process.
    * Relevant for subsequent modeling.

# Clustering Approaches

- Record linkage can be naturally seen as a clustering problem.
    * Supervised and unsupervised approches.
    
\vspace{2mm}

- Records representing the same individual are clustered to a latent entity producing a partition of the data.
    * Steorts, R., Hall, R., and Fienberg, S.E. (2016). A Bayesian Approach to Graphical Record Linkage and De-duplication, Journal of the American Statistical Association, 111:516 (1660-1672).
    \vspace{1mm}
    * Sadinle, M. (2014). Detecting duplicates in a homicide registry using a
bayesian partitioning approach. The Annals of Applied Statistics, Vol. 8, No. 4, 2404–2434

# Record Linkage and Clustering 

Which records correspond to the same person?
\begin{figure}
\includegraphics[scale=0.35,keepaspectratio]{figures/DukeRL}
\end{figure}

# Record Linkage and Clustering 

Each entity is associated with one or more records and the
goal is to recover the latent entities (clusters).
\begin{figure}
\includegraphics[scale=0.3,keepaspectratio]{figures/DukeRLlatent}
\end{figure}

# Graphical Record Linkage

Graphical model representation of  \textcolor{blue}{Steorts et al. (2016)}:
\begin{figure}
\includegraphics[scale=0.35,keepaspectratio]{figures/recordLinkage_graphicalModel}
\end{figure}
- $\Lambda_{ij}$ represents the linkage structure $\rightarrow$  \textcolor{blue}{uniform prior}.
<!--uniform over the set of {1,...,N}
such that each record is assumed to be equally likely a priori to correspond to any of the N latent individuals. Thus, it treats the records as if they are a random sample drawn with replacement.
% any two partitions with the same number of latent entities are equally probably a priori.--> 
- Requires information about the number of latent entities a priori and it is very informative.

# Partition-based Bayesian clustering models

Goal: cluster N data points $x_1,\cdots, x_N$ into $K$ clusters.

\vspace{1mm}

- Place a prior distribution over partitions of $[N] = \{1,\ldots,N\}$

\vspace{1mm}

- Let $C_N$ be a random partition of $[N]$

\vspace{1mm}

- $C_N$ represented by a set of cluster assignments $z_1,\ldots ,z_N$.

\vspace{1mm}

- The number of clusters $K$ does not need to be specified a priori $\rightarrow$
\textcolor{blue}{Non-parametric} latent variable approach.


# Record Linkage and Microclustering

\begin{figure}
\vspace*{-1.3cm}
\hspace*{7.5cm} \includegraphics[scale=0.24,keepaspectratio]{figures/HRDAG-Staging-Logo-2}
\end{figure}
\hspace{-1cm}
\begin{minipage}{4cm}
\vspace{-1cm}
%\includegraphics[height=5cm, width=3.5cm]{horner}
 \includegraphics[scale=0.45,keepaspectratio]{figures/syria-sizes}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}{7cm}
\vspace{-1.5cm}
- Enumeration of victims of killings in Syria merging four databases.
\vspace{1cm}

- The number of data points in each cluster should remain small even for large data sets $\rightarrow$ Large number of singletons and small clusters.
\end{minipage}

# Mixture Models

Other clustering tasks require models that assume cluster sizes
grow linearly with the size of the data set.

\begin{itemize}

\item Dirichlet process (DP)  $\Longrightarrow$
Chinese Restaurant Process (CRP)
\begin{figure}
\includegraphics[scale=0.6,keepaspectratio]{figures/chinesealpha}
\end{figure}
\item Carmona C., Nieto-Barajas L., Canale A. (2017), Model-based approach for household clustering with mixed scale variables \url{https://arxiv.org/abs/1612.00083} .
\end{itemize}

<!--The Ministry of Social Development in Mexico is in charge of creating and assigning social programmes targeting specific needs in the population for the improvement of quality of life. To better target the social programmes, the Ministry is aimed to find clusters of households with the same needs based on demographic characteristics as well as poverty conditions of the household.--> 

# Microclustering models

- Prior distributions on partitions that are suitable for the microclustering problem.
    * Zanella et al (2016). Flexible Models for Microclustering with Applications to Entity Resolution, Advances in Neural Information Processing Systems (NIPS), Vol. 29, pp 1417-1425.

\vspace{1mm}

- Scalable sampling algorithm in combination with blocking techniques.
    * Miller et al (2015). The Microclustering Problem: When the
Cluster Sizes Don't Grow with the Number of Data Points. NIPS Bayesian Nonparametrics: The Next Generation Workshop Series.

# `blink` package

`R` package that removes duplicate entries from multiple databses using the empirical Bayes graphical method:

```{r blink-install, eval=FALSE, echo=TRUE}
install.packages("blink")
```

- Formatting data for use with `blink`
- Tuning parameters
- Running the Gibbs sampler (estimate model parameters)
- Output


# `RLdata500` data

We will continue with the  `RLdata500` dataset in the `RecordLinkage` package consisting of $500$ records with 10% duplication.

```{r data, echo=TRUE}
library(blink) # load blink library
library(RecordLinkage) # load data library
data("RLdata500") # load data
head(RLdata500) # take a look
```

# Formatting the data

```{r, echo=TRUE}
# categorical variables
X.c <- as.matrix(RLdata500[, c("by","bm","bd")]) 
p.c <- ncol(X.c) 

# string variables 
X.s <- as.matrix(RLdata500[, c("fname_c1", "lname_c1")]) 
p.s <- ncol(X.s) 
```

`X.c` and `X.s` include all files stacked on top of each other, for categorical and string variables respectively

```{r echo=TRUE}
# keep track of which rows of are in which files
file.num <- rep(c(1, 2, 3), c(200, 150, 150))
```

# Tuning parameters

## Hyperparameters

```{r, echo=TRUE}
# Subjective choices for distortion probability prior
# parameters of a Beta(a,b)
a <- 1
b <- 999
```

## Distortion

```{r, echo=TRUE}
# string distance function example
d <- function(s1, s2) {
  adist(s1, s2) # approximate string distance
}

# steepness parameter
c <- 1
```

# Running the Gibbs sampler

```{r, echo=TRUE, cache=TRUE, results='hide'}
lam.gs <- rl.gibbs(file.num = file.num, # file
                   X.s = X.s, X.c = X.c, # data 
                   num.gs = 100000, # iterations
                   a = a, b = b, # prior params
                   c = c, d = d, # distortion
                   M = 500) # max # latents
```

# Output

\scriptsize

```{r get-estimates-sizes, echo=TRUE, results="hide"}
# count how many unique latent individuals
size_est <- apply(lam.gs, 1, function(x) { 
  length(unique(x)) 
  })
```

```{r plots, fig.width=2.25, fig.height=2.5, fig.show='hold'}
theme_set(theme_bw(base_family = "serif", base_size = 15))

ggplot() +
  geom_line(aes(1:100000, size_est)) +
  xlab("Iteration") +
  ylab("Estimated size")

ggplot() +
  geom_histogram(aes(size_est[-(1:25000)])) +
  geom_vline(aes(xintercept=450), colour = "red") +
  geom_vline(aes(xintercept=mean(size_est[-(1:25000)])), lty=2) +
  xlab("Estimated size") +
  xlim(c(400, 500))
  
```

# Evaluation
\scriptsize
```{r, echo=TRUE, cache=TRUE}
# estimated pairwise links
est_links_pair <- pairwise(links(lam.gs[-(1:25000), ]))

# true pairwise links
true_links_pair <- pairwise(links(matrix(identity.RLdata500, nrow = 1)))

#comparison
comparison <- links.compare(est_links_pair, true_links_pair, counts.only = TRUE)

# precision
precision <- comparison$correct/(comparison$incorrect + comparison$correct)

# recall
recall <- comparison$correct/(comparison$correct + comparison$missing)


# results
c(precision, recall)
```


# Your turn



